---
title: "Explore Weather Trends"
author: "Shubham Kumar"
date: "03/23/2020"
output:
  md_document:
    variant: markdown_github
  pdf_document:
    toc: yes
    toc_depth: '2'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

# Tool Used : R and Rmd

# Getting the Data.

## Extract data from database
Download the global data as csv from the course site, using the command below.
```
SELECT * from global_data;
```
Cities data can be downloaded from the sql command below.
```
SELECT * from city_data;
```

However, if we need the data for specific cities (New Delhi, London, Barcelona and Berlin) we can provide where condition to the sql query. Since there are two London and Barcelona in different countries we will also specify country (United Kingdom and Spain).
```
SELECT * from city_data 
WHERE city = 'New Delhi' and country = 'India';

SELECT * from city_data
WHERE city = 'London' and country = 'United Kingdom';

SELECT * from city_data 
WHERE city = 'Barcelona' and country = 'Spain';

SELECT * from city_data 
WHERE city = 'Berlin' and country = 'Germany';
```

## Install R packages needed in the project.

```{r results=FALSE}
library(ggplot2)
library(tseries)
library(forecast)
```

## Load both the global and city data

```{r results = TRUE}
city_data <- read.csv("city_data.csv", header = TRUE)
global_data <- read.csv("global_data.csv", header = TRUE)

```

## Loading cities data (New Delhi, London, Barcelona and Berlin)

```{r results = TRUE}
city_data_new_delhi <- subset(city_data, city_data$city == 'New Delhi' & city_data$country == 'India')
city_data_london <- subset(city_data, city_data$city == 'London' & city_data$country == 'United Kingdom')
city_data_barcelona <- subset(city_data, city_data$city == 'Barcelona' & city_data$country == 'Spain')
city_data_berlin <- subset(city_data, city_data$city == 'Berlin' & city_data$country == 'Germany')


```

## Removing data with null values
```{r results = TRUE}
city_data_new_delhi <- subset(city_data_new_delhi, !is.na(city_data_new_delhi$avg_temp))
city_data_london <- subset(city_data_london, !is.na(city_data_london$avg_temp))
city_data_barcelona <- subset(city_data_barcelona, !is.na(city_data_barcelona$avg_temp))
city_data_berlin <- subset(city_data_berlin, !is.na(city_data_berlin$avg_temp))

```

## Taking only neccessary Data
```{r results = TRUE}
city_data_new_delhi <- city_data_new_delhi[c("year", "avg_temp")]
city_data_london <- city_data_london[c("year", "avg_temp")]
city_data_barcelona <- city_data_barcelona[c("year", "avg_temp")]
city_data_berlin <- city_data_berlin[c("year", "avg_temp")]
```

The last 4 commands above are used to clean the city data by removing the column that are not needed (the city name and country name).

# Exploring the Data.

## Summary of Global Data
```{r results=TRUE}
summary(global_data)
```

## Summary of New Delhi's Data
```{r results=TRUE}
summary(city_data_new_delhi)
```

## Summary of Londons's Data
```{r results=TRUE}
summary(city_data_london)
```

## Summary of Barcelona's Data
```{r results=TRUE}
summary(city_data_barcelona)
```


## Summary of Berlin's Data
```{r results=TRUE}
summary(city_data_berlin)
```

Now try to create a moving average plot of all data with size of 7.

#### Moving Average : 
Moving averages are used to smooth out data to make it easier to observe long term trends and not get lost in daily fluctuations. 

**The moving average is calculated by adding data over a certain period and dividing the sum by the total number of periods.**

### Formula for finding moving average : 
```{r}
m7 <- function(arr, n = 7){
  res = arr
  for(i in n:length(arr)){
    res[i] = mean(arr[(i-n):i])
  }
  res
}

```

**Another Way to find moving average is to : **

Create a repetition of 7 for taking moving average of 7 years and filter avg_temp with it. **(STACKOVERFLOW)**. Using this method as this is faster than using self made function.

```{r}
m7 <- rep(1/7,7)

```

## Creating moving average temperature for global data

```{r results=TRUE}
global_data$moving_avg_temp <- filter(global_data$avg_temp, m7, sides = 2)

```

## Creating moving average temperature for all cities data

```{r results=TRUE}
city_data_new_delhi$moving_avg_temp <- filter(city_data_new_delhi$avg_temp, m7, sides=2)
city_data_london$moving_avg_temp <- filter(city_data_london$avg_temp, m7, sides=2)
city_data_barcelona$moving_avg_temp <- filter(city_data_barcelona$avg_temp, m7, sides=2)
city_data_berlin$moving_avg_temp <- filter(city_data_berlin$avg_temp, m7, sides=2)

```


## Creating moving average plot for Global data
```{r warning=FALSE, results=TRUE}
head(global_data)

gplot1 <- ggplot(data = global_data, aes(global_data$year, global_data$moving_avg_temp))
gplot1 <- gplot1 + geom_line(size = 1, col = "darkcyan") + labs(title = "Moving Line Plot of Global Data", x = "Year", y = "Moving Average Temperature in Degree Celcius")
gplot1 <- gplot1 + geom_smooth(method = "lm", se = FALSE, col = 'red')
gplot1
```

## Creating moving average plot for New Delhi's data
```{r warning=FALSE, results=TRUE}
head(city_data_new_delhi)

gplot2 <- ggplot(data = city_data_new_delhi, aes(city_data_new_delhi$year, city_data_new_delhi$moving_avg_temp))
gplot2 <- gplot2 + geom_line(size = 1, col = "orange4") + labs(title = "Moving Line Plot of New Delhi Data", x = "Year", y = "Moving Average Temperature in Degree Celcius")
gplot2 <- gplot2 + geom_smooth(method = "lm", se = FALSE, col = 'red')
gplot2

```

## Creating moving average plot for new london's data
```{r warning=FALSE, results=TRUE}
head(city_data_london)

gplot3 <- ggplot(data = city_data_london, aes(city_data_london$year, city_data_london$moving_avg_temp, color = "purple"))
gplot3 <- gplot3 + geom_line(size = 1, col = "green4") + labs(title = "Moving Line Plot of London Data", x = "Year", y = "Moving Average Temperature in Degree Celcius")
gplot3 <- gplot3 + geom_smooth(method = "lm", se = FALSE, col = 'red')
gplot3

```

## Creating moving average plot for new Barcelona's data
```{r warning=FALSE, results=TRUE}
head(city_data_barcelona)

gplot4 <- ggplot(data = city_data_barcelona, aes(city_data_barcelona$year, city_data_barcelona$moving_avg_temp))
gplot4 <- gplot4 + geom_line(size = 1, col = "darkblue") + labs(title = "Moving Line Plot of Barcelona Data", x = "Year", y = "Moving Average Temperature in Degree Celcius")
gplot4 <- gplot4 + geom_smooth(method = "lm", se = FALSE, col = 'red')
gplot4

```

## Creating moving average plot for berlin's data
```{r warning=FALSE, results=TRUE}
head(city_data_berlin)

gplot5 <- ggplot(data = city_data_berlin, aes(city_data_berlin$year, city_data_berlin$moving_avg_temp))
gplot5 <- gplot5 + geom_line(size = 1, col = "purple3") + labs(title = "Moving Line Plot of Berlin Data", x = "Year", y = "Moving Average Temperature in Degree Celcius")
gplot5 <- gplot5 + geom_smooth(method = "lm", se = FALSE, col = 'red')
gplot5

```

# Observations

As we can see from above, cities and global temperature has tendency to increase throughout the year.

**My City (New Delhi) is hotter than any other countries in analysis but this is because it lies in tropical area.**

## Is the increase the same for city and global?

Lets calculate the difference between cities temperature Data and global Data.
```{r}
diff1 <- city_data_new_delhi$avg_temp - global_data$avg_temp
diff2 <- city_data_london$avg_temp - global_data$avg_temp
diff3 <- city_data_barcelona$avg_temp - global_data$avg_temp
diff4 <- city_data_berlin$avg_temp - global_data$avg_temp
```

Differnce between average temperatures of Global and New Delhi.
```{r results=TRUE}
summary(diff1)
```

Differnce between average temperatures of Global and London.
```{r results=TRUE}
summary(diff2)
```

Differnce between average temperatures of Global and Barcelona.
```{r results=TRUE}
summary(diff3)
```

Differnce between average temperatures of Global and Berlin.
```{r results=TRUE}
summary(diff4)

```

**From above summary, we can conclude that there's is a high mean difference between global data and New Delhi (16 degrees) and a minor difference between barcelona and global data. However, there's almost 0 difference between the temperatures at london and berlin.**

**This suggests that cities with warm weather are getting hotter while the cities with cool weather has no great change meaning they are increasing consistently with global temperatures but temperatures at warmer cities (New Delhi and Barcelona) are increasing faster than global temperature.**

## New Delhi and Global Data

Plotting the data of New Delhi and Global, it should be a nearly straight line if the increase is the same.


```{r results=TRUE}
plot(global_data$year, diff1, type="l", xlab="Year", ylab="Difference between New Delhi and Global temperature", col = 'blue')
```

**From the plot above, it is hard to tell whether the data is actually increasing or decreasing between years 1750 to 1950. However we can see that the difference lies around 16 degrees each year and trend is continuosly decreasing after the year 1950.**

Let's try to smoothen the data using moving average to visualize the differences better.
```{r results=TRUE}
diff_moving_avg_temp <- filter(diff1, m7, sides=2)
plot(global_data$year, diff_moving_avg_temp, type="l", xlab="Year", ylab="Difference between New Delhi and Global temperature",col = 'blue')
```

**From here, we can see that in the last few decades, the difference has tendency to decrease.**

It means global temperature has increased by more than city's since the difference is becoming smaller.

## Does global temperature increase faster than city temperature?

We will be using difference of 10 years between temperatures to compare the temperature of the year with the temperature 10 years ago and compare trend of global data with cities data.

```{r results=TRUE}
plot(diff(global_data$avg_temp, differences= 10), type="l", col = "blue", lwd = 3, xlab = "", ylab = "")
lines(diff(city_data_new_delhi$avg_temp, differences=10), col="red", lwd = 3)
```

**The red plot marks the new delhi's temperature difference, which is similar to the global temperature. Thus, we can roughly conclude that global temperature increase as same pace as the city temperature.**

```{r results=TRUE}
plot(diff(global_data$avg_temp, differences= 10), type="l", col = "blue", lwd = 3, xlab = "", ylab = "")
lines(diff(city_data_london$avg_temp, differences=10), col="red", lwd = 3)
```

**The red plot marks the london's temperature difference, which is higher to the global temperature. Thus, we can roughly conclude that london's temperature is increasing faster than global temperature.**

```{r results=TRUE}
plot(diff(global_data$avg_temp, differences= 10), type="l", col = "blue", lwd = 3, xlab = "", ylab = "")
lines(diff(city_data_barcelona$avg_temp, differences=10), col="red", lwd = 3)
```

**The red plot marks the barcelona's temperature difference, which is similar to the global temperature. Thus, we can roughly conclude that barcelona's temperature is increasing at same rate as the global temperature.**

```{r results=TRUE}
plot(diff(global_data$avg_temp, differences= 10), type="l", col = "blue", lwd = 3, xlab = "", ylab = "")
lines(diff(city_data_berlin$avg_temp, differences=10), col="red", lwd = 3)
```

**The red plot marks the berlin's temperature difference, which is a lot higher to the global temperature. Thus, we can roughly conclude that berlin's temperature is increasing at a lot faster than the global temperature.**


# To conclude, there is a clear overall uptrend visible, what means, that the average global temperature is increasing, with an also increasing tempo.

The cities New Delhi(India), Berlin(Germany), London(United Kingdom) and Barcelona(Spain) got compared to the global data.

### Clearly, Global temperatures are increasing with similar trend to other cities.

**Looking at the difference between temperatures of New Delhi and Global Data, it is almost 16 and constant for a long peroid but decreases after 1950 which means the global temperature is increasing at higher pace as there's a sharp decline in difference.**

**Barcelona and New Delhi has almost same maximum and mean temperatures but there is a lot of difference between there minimum temperatures as they both are at the same tropical zone, suggesting barcelona has almost similar temperature.**